Location and Scale-Invariant Power Transformations for Transforming Data to Normality

### 2025-04-09 manuscript_rmarkdown/manuscript.tex
Changed text to definitions.

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  a4paper,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{amsmath}
\usepackage{booktabs}
\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator{\sgn}{sgn}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Location and Scale-Invariant Power Transformations for Transforming Data to Normality},
  pdfauthor={Alex Zwanenburg, Steffen Löck},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Location and Scale-Invariant Power Transformations for
Transforming Data to Normality}
\author{Alex Zwanenburg, Steffen Löck}
\date{2024-10-17}

@@

Many statistical and some machine learning methods assume normality of
the underlying data, e.g.~analysis of variance and linear discriminant
analysis. However, numerical features in datasets may strongly deviate
from normal distributions, e.g.~by being skewed. Power transformations
aim to stabilise variance and improve normality of such features
(Bartlett 1947; John W. Tukey 1957). The two most commonly used
transformations are that of Box and Cox (1964) and Yeo and Johnson
(2000). The Box-Cox transformation of a feature value \(x_i\) of feature
\(\mathbf{X}=\left\{x_1, x_2, \ldots, x_n \right\}\) under the
transformation parameter \(\lambda\) is defined as:

\begin{equation}
\label{eqn:box-cox-original}
\phi_{\text{BC}}^\lambda (x_i) = 
\begin{cases}
\left(x_i^\lambda - 1 \right) / \lambda & \text{if } \lambda \neq 0\\
\log(x_i) & \text{if } \lambda = 0
\end{cases}
\end{equation}

One limitation of the Box-Cox transformation is that it is only defined
for \(x_i > 0\). In contrast, the Yeo-Johnson transformation under the
transformation parameter \(\lambda\) is defined for any
\(x_i \in \mathbb{R}\):

\begin{equation}
\label{eqn:yeo-johnson-original}
\phi_{\text{YJ}}^\lambda (x_i) = 
\begin{cases}
\left( \left( 1 + x_i \right)^\lambda - 1\right) / \lambda & \text{if } \lambda \neq 0 \text{ and } x_i \geq 0\\
\log(1 + x_i) & \text{if } \lambda = 0 \text{ and } x_i \geq 0\\
-\left( \left( 1 - x_i\right)^{2 - \lambda} - 1 \right) / \left(2 - \lambda \right) & \text{if } \lambda \neq 2 \text{ and } x_i < 0\\
-\log(1 - x_i) & \text{if } \lambda = 2 \text{ and } x_i < 0
\end{cases}
\end{equation}

@@

Box-Cox and Yeo-Johnson transformations are modified by introducing
shift parameter \(x_0\) and scale parameter \(s\) into equations
\ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}. The
location- and scale-invariant Box-Cox transformation of a feature value
\(x_i\) of feature \(\mathbf{X}\) under transformation parameter
\(\lambda\), shift parameter \(x_0\) and scale parameter \(s\) is then:

\begin{equation}
\label{eqn:box-cox-invariant}
\phi_{\text{BC}}^{\lambda, x_0, s} (x_i) = 
\begin{cases}
\left( \left(\frac{x_i - x_0}{s} \right)^\lambda - 1 \right) / \lambda & \text{if } \lambda \neq 0\\
\log\left[\frac{x_i - x_0}{s}\right] & \text{if } \lambda = 0
\end{cases}
\end{equation}

where \(x_i - x_0 > 0\). Likewise, the location- and scale-invariant
Yeo-Johnson transformation of a feature value \(x_i\) under
transformation parameter \(\lambda\), shift parameter \(x_0\) and scale
parameter \(s\) is:

\begin{equation}
\label{eqn:yeo-johnson-invariant}
\phi_{\text{YJ}}^{\lambda, x_0, s} (x_i) = 
\begin{cases}
\left( \left( 1 + \frac{x_i - x_0}{s}\right)^\lambda - 1\right) / \lambda & \text{if } \lambda \neq 0 \text{ and } x_i - x_0 \geq 0\\
\log\left[1 + \frac{x_i - x_0}{s}\right] & \text{if } \lambda = 0 \text{ and } x_i - x_0 \geq 0\\
-\left( \left( 1 - \frac{x_i - x_0}{s}\right)^{2 - \lambda} - 1 \right) / \left(2 - \lambda \right) & \text{if } \lambda \neq 2 \text{ and } x_i - x_0 < 0\\
-\log\left[1 - \frac{x_i - x_0}{s}\right] & \text{if } \lambda = 2 \text{ and } x_i - x_0 < 0
\end{cases}
\end{equation}

@@

\begin{equation}
\label{eqn:box-cox-invariant-log-likelihood}
\begin{split}
\mathcal{l}_{\text{BC}}^{\lambda, x_0, s} = & -\frac{n}{2} \log \left[2 \pi \sigma^2 \right] -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi_{BC}^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& -n \lambda \log s + \left( \lambda - 1 \right) \sum_{i=1}^n \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}

subject to \(x_i - x_0 > 0\). \(\mu\) and \(\sigma^2\) are the mean and
variance of the Box-Cox transformed feature
\(\phi_{\text{BC}}^{\lambda, x_0, s} (\mathbf{X})\), respectively.
Similarly, the location- and scale-invariant Yeo-Johnson log-likelihood
function is:

\begin{equation}
\label{eqn:yeo-johnson-invariant-log-likelihood}
\begin{split}
\mathcal{l}_{\text{YJ}}^{\lambda, x_0, s} = & -\frac{n}{2} \log\left[2 \pi \sigma^2\right] -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi_{YJ}^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& - n \log s + (\lambda - 1) \sum_{i=1}^n \sgn(x_i - x_0) \log \left[1 + \frac{|x_i - x_0|}{s} \right]
\end{split}
\end{equation}

where \(\mu\) and \(\sigma^2\) are the mean and variance of the
Yeo-Johnson transformed feature
\(\phi_{\text{YJ}}^{\lambda, x_0, s} (\mathbf{X})\), respectively.

@@

\begin{equation}
f_{\beta}(x) = \frac{\beta}{2\Gamma\left(1 / \beta \right)} e^{-\left| x \right|^\beta}
\end{equation}

Here, \(\Gamma\) is the gamma function, and \(\beta\) is a strictly
positive shape parameter. For \(\beta = 1\), the probability density
function describes a Laplace distribution. A normal distribution is
found for \(\beta=2\), and for large \(\beta\), the distribution
approaches a uniform distribution. We will refrain from introducing
scale and location parameters here directly.

@@

\begin{equation}
f_{\alpha}(x; \mu, \sigma, \beta) = \frac{2 \alpha \left(1 - \alpha\right)}{\sigma}
\begin{cases}
f_{\beta}\left( \left(1 - \alpha \right) \frac{\left| x - \mu \right|}{\sigma} \right) & \text{, } x \leq \mu \\
f_{\beta}\left( \alpha \frac{\left| x - \mu \right|}{\sigma} \right) & \text{, } x > \mu
\end{cases}
\end{equation}

Here, \(\alpha \in (0,1)\) is a skewness parameter. \(\alpha > 0.5\)
creates a distribution with a negative skew, i.e.~a left-skewed
distribution. A right-skewed distribution is created for
\(\alpha < 0.5\). \(\mu \in \mathcal{R}\) and \(\sigma \in (0, \infty)\)
are location and scale parameters, respectively. \(f_{\alpha}\) thus
describes the probability density function of an asymmetric generalised
normal distribution, which we will refer to here and parametrise as
\(\mathcal{AGN}\left(\mu, \sigma, \alpha, \beta \right)\).

We require a quantile function (or an approximation thereof) to draw
random values from an asymmetric generalised normal distribution using
inverse transform sampling. Gijbels et al.~derived the following
quantile function \(F_{\alpha}^{-1}(p)\):

\begin{equation}
F_{\alpha}^{-1}(p; \mu, \sigma, \beta) =
\begin{cases}
\mu + \frac{\sigma}{1 - \alpha} F_{\beta}^{-1} \left( \frac{p}{2 \alpha}\right) & \text{, } p \leq \alpha \\
\mu + \frac{\sigma}{\alpha} F_{\beta}^{-1} \left( \frac{1 + p - 2 \alpha}{2 \left(1 - \alpha \right)} \right) & \text{, } p > \alpha
\end{cases}
\end{equation}

The quantile function for the asymmetric generalised normal distribution
\(F_{\alpha}^{-1}\) thus incorporates the quantile function
\(F_{\beta}^{-1}\) of the symmetric generalised normal distribution.
\(F_{\beta}^{-1}\) was derived by Griffin to be (Griffin 2018):

\begin{equation}
F_{\beta}^{-1}(p) = \sgn\left(p - 0.5 \right) F_{\Gamma}^{-1}\left(2 \left|p - 0.5 \right|; 1 / \beta \right)
\end{equation}

Here, \(F_{\Gamma}^{-1}\) is the quantile function of the gamma
distribution with shape \(1 / \beta\), which can be numerically
approximated.

\subsection{Empirical central normality
test}\label{empirical-central-normality-test}

Power transformations aim to transform features to a normal
distribution. However, this may not always be successful or possible.
Deviations from normality can be detected by normality tests, such as
the Shapiro-Wilk test (Shapiro and Wilk 1965). In practice, normality
tests may be too stringent with large sample sizes, outliers, or both.
Here we develop an empirical test for central normality. The null
hypothesis \(H_0\) is that the distribution is centrally normal. The
alternative hypothesis \(H_1\) is that the distribution is not centrally
normal.

Let central normality be defined as the normality of central portion
\(\kappa\) of the data,
i.e.~\(\mathbf{X}_{\text{central}} = \left\{x_i \in \mathbf{X} \, | \,  \frac{1-\kappa}{2} \leq  p_i \leq \frac{1 + \kappa}{2}\right\}\),
with \(p_i\) probabilities of the empirical distribution, as previously.
We then compute the residual errors between the z-scores of the
transformed feature \(\phi^{\lambda, x_0, s} (\mathbf{X})\) and the
expected z-scores from a standard normal distribution:
\(r_i =\left| \left( \phi^{\lambda, x_0, s}(x_i) - \mu_M)\right) / \sigma_M - F^{-1}_{\mathcal{N}}(p_i) \right|\),
with \(\mu_M\) and \(\sigma_M\) robust Huber M-estimates of location and
scale of the transformed feature \(\phi^{\lambda, x_0, s} (\mathbf{X})\)
(Huber 1981). The set of residual errors for the central portion of the
data is then
\(\mathbf{R}_{\text{central}} = \left\{ r_i \in \left\{ r_1, r_2, \ldots, r_n\right\} \, | \,  \frac{1-\kappa}{2} \leq  p_i \leq \frac{1 + \kappa}{2}\right\}\).

The test statistic \(\tau_{\text{ecn}}\) is then defined as:

\begin{equation}
\tau_{\text{ecn}} = \frac{\sum_{i=1}^{N} r_i \left[r_i \in \mathbf{R}_{\text{central}}\right]}{\sum_{i=1}^N \left[r_i \in \mathbf{R}_{\text{central}}\right]} 
\end{equation}

Here \([\quad]\) denotes an Iverson bracket. The test statistic is equal
to the mean of the residual errors of the central portion of the data.

To use the test statistic, the central portion of the data needs to be
defined and the Type 1 error rates determined. We will do so in the
\hyperref[simulation]{Simulation} section.

@@

{\centering \includegraphics{manuscript_files/figure-latex/optimised-weighting-function-parameters-1}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  a4paper,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{booktabs}
\newtheorem*{definition}{Definition}
\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator{\sgn}{sgn}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Location and Scale-Invariant Power Transformations for Transforming Data to Normality},
  pdfauthor={Alex Zwanenburg, Steffen Löck},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Location and Scale-Invariant Power Transformations for
Transforming Data to Normality}
\author{Alex Zwanenburg, Steffen Löck}
\date{2025-04-09}

@@

Many statistical and some machine learning methods assume normality of
the underlying data, e.g.~analysis of variance and linear discriminant
analysis. However, numerical features in datasets may strongly deviate
from normal distributions, e.g.~by being skewed. Power transformations
aim to stabilise variance and improve normality of such features
(Bartlett 1947; John W. Tukey 1957). The two most commonly used
transformations are that of Box and Cox (1964) and Yeo and Johnson
(2000).

The Box-Cox transformation is defined as:

\begin{definition}[Box-Cox power transformation]
Let $\mathbf{X} = \left\{x_1, x_2, \ldots, x_n | x_i > 0 \right\}$ be a finite sequence of length $n > 0$.
Let $\lambda \in \mathcal{R}$ be a transformation parameter.
Then the Box-Cox power transformation of element $x_i$ is defined as:
\begin{equation}
\label{eqn:box-cox-original}
\phi_{\text{BC}}^\lambda (x_i) = 
\begin{cases}
\left(x_i^\lambda - 1 \right) / \lambda & \text{if } \lambda \neq 0\\
\log(x_i) & \text{if } \lambda = 0
\end{cases}
\end{equation}
\end{definition}

One limitation of the Box-Cox transformation is that it is only defined
for \(x_i > 0\). In contrast, the Yeo-Johnson transformation under the
transformation parameter \(\lambda\) is defined for any
\(x_i \in \mathbb{R}\), as follows:

\begin{definition}[Yeo-Johnson power transformation]
Let $\mathbf{X} = \left\{x_1, x_2, \ldots, x_n | x_i \in \mathbb{R} \right\}$ be a finite sequence of length $n > 0$.
Let $\lambda \in \mathcal{R}$ be a transformation parameter.
\begin{equation}
\label{eqn:yeo-johnson-original}
\phi_{\text{YJ}}^\lambda (x_i) = 
\begin{cases}
\left( \left( 1 + x_i \right)^\lambda - 1\right) / \lambda & \text{if } \lambda \neq 0 \text{ and } x_i \geq 0\\
\log(1 + x_i) & \text{if } \lambda = 0 \text{ and } x_i \geq 0\\
-\left( \left( 1 - x_i\right)^{2 - \lambda} - 1 \right) / \left(2 - \lambda \right) & \text{if } \lambda \neq 2 \text{ and } x_i < 0\\
-\log(1 - x_i) & \text{if } \lambda = 2 \text{ and } x_i < 0
\end{cases}
\end{equation}
\end{definition}

@@

Box-Cox and Yeo-Johnson transformations are modified by introducing
shift parameter \(x_0\) and scale parameter \(s\) into equations
\ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}. The
location- and scale-invariant Box-Cox transformation of a feature value
\(x_i\) of feature \(\mathbf{X}\) under transformation parameter
\(\lambda\), shift parameter \(x_0\) and scale parameter \(s\) is then
defined as:

\begin{definition}[Location- and scale- invariant Box-Cox power transformation]
Let $\lambda \in \mathcal{R}$ be a transformation parameter.
Let $x_0 \in \mathcal{R}$ and $s > 0$ be location and scale parameters.
Let $\mathbf{X} = \left\{x_1, x_2, \ldots, x_n | x_i - x_0 > 0 \right\}$ be a finite sequence of length $n > 0$.

Then the location- and scale-invariant Box-Cox power transformation of element $x_i$ is defined as:

\begin{equation}
\label{eqn:box-cox-invariant}
\phi_{\text{BC}}^{\lambda, x_0, s} (x_i) = 
\begin{cases}
\left( \left(\frac{x_i - x_0}{s} \right)^\lambda - 1 \right) / \lambda & \text{if } \lambda \neq 0\\
\log\left[\frac{x_i - x_0}{s}\right] & \text{if } \lambda = 0
\end{cases}
\end{equation}
\end{definition}

Likewise, the location- and scale-invariant Yeo-Johnson transformation
of a feature value \(x_i\) under transformation parameter \(\lambda\),
shift parameter \(x_0\) and scale parameter \(s\) is defined as:

\begin{definition}[Location- and scale- invariant Yeo-Johnson power transformation]
Let $\lambda \in \mathcal{R}$ be a transformation parameter.
Let $x_0 \in \mathcal{R}$ and $s > 0$ be location and scale parameters.
Let $\mathbf{X} = \left\{x_1, x_2, \ldots, x_n | x_i \in \mathcal{R} \right\}$ be a finite sequence of length $n > 0$.

Then the location- and scale-invariant Yeo-Johnson power transformation of element $x_i$ is defined as:

\begin{equation}
\label{eqn:yeo-johnson-invariant}
\phi_{\text{YJ}}^{\lambda, x_0, s} (x_i) = 
\begin{cases}
\left( \left( 1 + \frac{x_i - x_0}{s}\right)^\lambda - 1\right) / \lambda & \text{if } \lambda \neq 0 \text{ and } x_i - x_0 \geq 0\\
\log\left[1 + \frac{x_i - x_0}{s}\right] & \text{if } \lambda = 0 \text{ and } x_i - x_0 \geq 0\\
-\left( \left( 1 - \frac{x_i - x_0}{s}\right)^{2 - \lambda} - 1 \right) / \left(2 - \lambda \right) & \text{if } \lambda \neq 2 \text{ and } x_i - x_0 < 0\\
-\log\left[1 - \frac{x_i - x_0}{s}\right] & \text{if } \lambda = 2 \text{ and } x_i - x_0 < 0
\end{cases}
\end{equation}
\end{definition}

@@

\begin{definition} [Log-likelihood function of the location- and scale- invariant Box-Cox power transformation]
Let $\mathbf{X}$, $\lambda$, $x_0$, and $s$ be defined as earlier (Eqn. \ref{eqn:box-cox-invariant}).
Let $\mu$ and $\sigma^2$ be the mean and variance of the transformed sequence $\phi_{\text{BC}}^{\lambda, x_0, s} (\mathbf{X})$, respectively.

Then the log-likelihood function of the location- and scale- invariant Box-Cox power transformation is defined as:
\begin{equation}
\label{eqn:box-cox-invariant-log-likelihood}
\begin{split}
\mathcal{l}_{\text{BC}}^{\lambda, x_0, s} = & -\frac{n}{2} \log \left[2 \pi \sigma^2 \right] -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi_{BC}^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& -n \lambda \log s + \left( \lambda - 1 \right) \sum_{i=1}^n \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}
\end{definition}

Similarly, the location- and scale-invariant Yeo-Johnson log-likelihood
function is:

\begin{definition}
Let $\mathbf{X}$, $\lambda$, $x_0$, and $s$ be defined as earlier (Eqn. \ref{eqn:yeo-johnson-invariant}).
Let $\mu$ and $\sigma^2$ be the mean and variance of the transformed sequence $\phi_{\text{YJ}}^{\lambda, x_0, s} (\mathbf{X})$, respectively.

Then the log-likelihood function of the location- and scale- invariant Yeo-Johnson power transformation is defined as:
\begin{equation}
\label{eqn:yeo-johnson-invariant-log-likelihood}
\begin{split}
\mathcal{l}_{\text{YJ}}^{\lambda, x_0, s} = & -\frac{n}{2} \log\left[2 \pi \sigma^2\right] -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi_{YJ}^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& - n \log s + (\lambda - 1) \sum_{i=1}^n \mathop{\mathrm{sgn}}(x_i - x_0) \log \left[1 + \frac{|x_i - x_0|}{s} \right]
\end{split}
\end{equation}
\end{definition}

@@

\begin{definition}[Standard generalised normal distribution probability density function]
Let $x \in \mathcal{R}$.
Let $\beta > 0$ be a shape parameter.
Let $\Gamma$ be the gamma function.

Then the standard generalised normal distribution probability density function, without scale and location parameters, is:
\begin{equation}
f_{\beta}(x) = \frac{\beta}{2\Gamma\left(1 / \beta \right)} e^{-\left| x \right|^\beta}
\end{equation}
\end{definition}

For \(\beta = 1\), the probability density function describes a Laplace
distribution. A normal distribution is found for \(\beta=2\), and for
large \(\beta\), the distribution approaches a uniform distribution.

@@

\begin{definition}[Asymmetric generalised normal distribution probability density function]
Let $\alpha \in (0,1)$ be a skewness parameter.
Let $\mu \in \mathcal{R}$ be a location parameter.
Let $\sigma > 0$ be a shape parameter.

Then the probability density function of the asymmetric generalised normal distribution is:
\begin{equation}
f_{\alpha}(x; \mu, \sigma, \beta) = \frac{2 \alpha \left(1 - \alpha\right)}{\sigma}
\begin{cases}
f_{\beta}\left( \left(1 - \alpha \right) \frac{\left| x - \mu \right|}{\sigma} \right) & \text{, } x \leq \mu \\
f_{\beta}\left( \alpha \frac{\left| x - \mu \right|}{\sigma} \right) & \text{, } x > \mu
\end{cases}
\end{equation}
\end{definition}

\(\alpha > 0.5\) creates a distribution with a negative skew, i.e.~a
left-skewed distribution. A right-skewed distribution is created for
\(\alpha < 0.5\). \(f_{\alpha}\) thus describes the probability density
function of an asymmetric generalised normal distribution, which we will
refer to here and parametrise as
\(\mathcal{AGN}\left(\mu, \sigma, \alpha, \beta \right)\).

We require a quantile function (or an approximation thereof) to draw
random values from an asymmetric generalised normal distribution using
inverse transform sampling. Gijbels et al.~derived the quantile function
\(F_{\alpha}^{-1}(p)\), which incorporates the quantile function of the
symmetric generalised normal distribution derived by Griffin (Griffin
2018):

\begin{definition}[Asymmetric generalised normal distribution quantile function]
Let $p \in \left[0, 1\right]$ be a probability.
Let  $F_{\Gamma}^{-1}$ is the quantile function of the gamma distribution with shape $1 / \beta$, which can be numerically approximated. 
Let $F_{\beta}^{-1}$ be the quantile function of the generalised normal distribution:
\begin{equation}
F_{\beta}^{-1}(p) = \mathop{\mathrm{sgn}}\left(p - 0.5 \right) F_{\Gamma}^{-1}\left(2 \left|p - 0.5 \right|; 1 / \beta \right)
\end{equation}

Then the quantile function of the asymmetric generalised normal distribution is:
\begin{equation}
F_{\alpha}^{-1}(p; \mu, \sigma, \beta) =
\begin{cases}
\mu + \frac{\sigma}{1 - \alpha} F_{\beta}^{-1} \left( \frac{p}{2 \alpha}\right) & \text{, } p \leq \alpha \\
\mu + \frac{\sigma}{\alpha} F_{\beta}^{-1} \left( \frac{1 + p - 2 \alpha}{2 \left(1 - \alpha \right)} \right) & \text{, } p > \alpha
\end{cases}
\end{equation}
\end{definition}

\subsection{Central normality test}\label{central-normality-test}

Power transformations aim to transform features to a normal
distribution. However, this may not always be successful or possible.
Deviations from normality can be detected by normality tests, such as
the Shapiro-Wilk test (Shapiro and Wilk 1965). In practice, normality
tests may be too stringent with large sample sizes, outliers, or both.
Here we develop a test for central normality.

\begin{definition}[Central portion of a sequence]
Let $\mathbf{X} = \left\{x_1, x_2, \ldots, x_n | x_i \in \mathbb{R}\right\}$ be a finite sequence of length $n > 0$,
ordered so that $x_1 \leq x_2 \leq \ldots \leq x_n$.
Let $p_i = \frac{i - 1/3}{n + 1/3}$, with $i = 1, 2, \ldots n$ be the percentile value
corresponding to each element in $\mathbf{X}$. For elements with tied values in $\mathbf{X}$,
percentile values are replaced by the average in their group, i.e.
if $x_j = x_{j+1} = \ldots = x_{j+m}$, then 
$p_j' = p_{j+1}' = \ldots = p_{j+m}' = 1/(m+1)\sum_j^{j+m}p_j$.
Furthermore, let central portion $\kappa \in (0,1)$.

Then the central portion of $\mathbf{X}$ is 
$\mathbf{X}_{\text{c}} = \left\{x_i \in \mathbf{X} \, | \,  \frac{1-\kappa}{2} \leq  p_i \leq \frac{1 + \kappa}{2}\right\}$.

\end{definition}

\begin{definition}[Residual errors of the central portion of a sequence]

Let the residual error for each element of $\mathbf{X}$ be $r_i =\left| \frac{x_i - \mu_M}{\sigma_M} - F^{-1}_{\mathcal{N}}(p_i) \right|$, 
with $\mu_M$ and $\sigma_M$ robust Huber M-estimates of location and scale of $\mathbf{X}$ [@Huber1981-su], 
and $F^{-1}_{\mathcal{N}}$ the quantile function of the normal distribution $\mathcal{N}(0, 1)$.

Then, the set of residual errors of the central portion of $\mathbf{X}$ (i.e. $\mathbf{X}_{\text{c}}$) is
$\mathbf{R}_{\text{c}} = \left\{ r_i \in \left\{ r_1, r_2, \ldots, r_n\right\} \, | \,  \frac{1-\kappa}{2} \leq  p_i \leq \frac{1 + \kappa}{2}\right\}$.

\end{definition}

\begin{definition}[Central normality of a sequence]

The central portion of a sequence is normally distributed if the sum of residual
errors is equal to zero:
$\sum_{r_j \in \mathbf{R}_{\text{c}}} r_j = 0$

\end{definition}

In practice, a finite sequence sampled from a normal distribution
\(\mathcal{N}(\mu, \sigma)\) will have a non-zero sum of residual
errors.

\begin{definition}[Central normality test]
The null-hypothesis is defined as: $\mathcal{H}_0: \sum_{r_j \in \mathbf{R}_{\text{c}}}r_j = 0$

The alternative hypothesis is defined as: $\mathcal{H}_1: \sum_{r_j \in \mathbf{R}_{\text{c}}}r_j > 0$
\end{definition}

The null-hypothesis \(\mathcal{H}_0\) is that the central portion of
sequence \(\mathbf{X}\) is normally distributed, with the alternative
hypothesis \(\mathcal{H}_1\) that it is not normally distributed. To
test this hypothesis, a test statistic is computed and compared against
a critical value.

\begin{definition}[Central normality test statistic]
Let $m$ be the number of elements of $\mathbf{R}_{\text{c}}$

The test statistic is then $\tau_{n, \kappa} = \frac{1}{m}\sum_{r_j \in \mathbf{R}_{\text{c}}}r_j$.

\end{definition}

The null-hypothesis should be rejected at significance level \(\alpha\)
if \(\tau_{n, \kappa} \geq \tau_{\alpha, n, \kappa, \text{critical}}\).

The central portion of the data needs to be defined and the Type 1 error
rates determined to provide critical test statistics. We will do so in
the \hyperref[simulation]{Simulation} section.

@@

{\centering \includegraphics{manuscript_files/figure-latex/empirical-central-normality-kappa-1}
